Automatically proving equivalence by type-safe reflection

### 2016-01-19 paper-JFP-DTP/content/1-introduction.tex
Almost done. futur work to write
Final fix before second submission

\subsection{Motivating example: Verified Binary Arithmetic}

@@

Ring solvers are already implemented for various proof
assistants, including Coq~\cite{Coq2005} and Agda\footnote{\url{http://wiki.portal.chalmers.se/agda/\%22?n=Libraries.UsingTheRingSolver}}. 
In this paper, we describe a certified
implementation of an automatic prover for equalities in a
\emph{hierarchy} of algebraic
structures, including Monoid, Groups and Rings (all potentially commutative),
for the Idris language.

@@

        \item We develop a hierarchy of tactics where each tactic reuses the
        rewriting machinery from the structure immediately below in the hierarchy.
        For example,
        simplifying neutral elements is only implemented at the monoid level,
        and each level above monoid reuses it. Re-usability can be quite tricky
        to obtain when we want to reuse a prover from a structure which is less
        expressive. For example, reusing the monoid prover for building the
        group prover is not trivial, since we lose the possibility to express
        negations ($-x$) and subtractions ($x-y$). Some specific encodings
        will be presented in Section~\ref{sect:reuse} for solving this problem.

>>>>

\subsection{Motivating example: Verified Binary Arithmetic}
\label{sect:motivatingExample}

@@

Ring solvers are already implemented for various proof
assistants, including Coq~\cite{Coq2005} and Agda\footnote{\url{http://wiki.portal.chalmers.se/agda/\%22?n=Libraries.UsingTheRingSolver}}. 
In this paper, we describe a certified
implementation\footnote{The implementation of our hierarchy of tactics can be found online at \url{https://github.com/FranckS/RingIdris}} of an automatic prover for equalities in a
\emph{hierarchy} of algebraic
structures, including Monoid, Groups and Rings (all potentially commutative),
for the Idris language.

@@

        \item We develop a hierarchy of tactics where each tactic reuses the
        rewriting machinery from the structure immediately below in the hierarchy.
        For example,
        simplifying neutral elements is only implemented at the monoid level,
        and each level inheriting from it will reuse it. Re-usability can be quite tricky
        to obtain when we want to reuse a prover from a structure which is less
        expressive. For example, reusing the monoid prover for building the
        group prover is not trivial, since we lose the possibility to express
        negations ($-x$) and subtractions ($x-y$). Some specific encodings
        will be presented in Section~\ref{sect:reuse} for solving this problem.

### 2016-01-19 paper-JFP-DTP/content/4-detailsNormalisation.tex
corrections with Edwin advices on latex
Final fix before second submission

We now describe briefly how the normalisation functions work for some of the different algebraic structures, and we also explain how we've been able to reuse each prover for the more specialised structures. The construction of the desired proof is always done bit by bit, because every rewriting of the reflected term is immediately justified by a little accompanying proof which shows the equivalence of the previous and new index.
\begin{figure}[H]
\figrule
\begin{center}
\begin{lstlisting}
littleRewriting : (p:algebraic structure on c) $\rightarrow$ ($\Gamma$:Vect n c) 
       $\rightarrow$ {c1:c} $\rightarrow$ (Expr p $\Gamma$ c1) $\rightarrow$ (c2 ** (Expr p $\Gamma$ c2, c1 $\simeq$ c2))
littleRewriting p $\Gamma$ e = (new expression, justification)
\end{lstlisting}
\end{center}
\caption{General pattern of a function doing a part of the normalisation}
\figrule
\end{figure}

@@

\begin{figure}[H]
\figrule
\begin{center}
\begin{lstlisting}
assoc : (p:SemiGroup c) $\rightarrow$ ($\Gamma$:Vect n c) $\rightarrow$ {c1:c} 
        $\rightarrow$ (ExprSG p $\Gamma$ c1) $\rightarrow$ (c2 ** (ExprSG p $\Gamma$ c2, c1 $\simeq$ c2))
assoc p $\Gamma$ (PlusSG 
             (PlusSG ex (ConstSG _ _ const1)) 
             (PlusSG (ConstSG _ _ const2) ey)) =
    let (r_ihx ** (e_ihx, p_ihx)) = (assoc p $\Gamma$ ex) in
    let (r_ihy ** (e_ihy, p_ihy)) = (assoc p $\Gamma$ ey) in
    let (r_3 ** (e_3, p_3)) = magmaReduce (semiGroup_to_magma 
                             (PlusSG (ConstSG _ _ const1) (ConstSG _ _ const2))) in
    let e_3' = magma_to_semiGroup p e_3 in
        (_ ** ((PlusSG (PlusSG e_ihx e_3') e_ihy), ?Massoc1))
\end{lstlisting}
\end{center}
\caption{Computing with associativity in a Semi-Group, first pattern}
\figrule
\end{figure}
The treatment is applied recursively on the sub-expressions $ex$ and $ey$, which produces the new expressions $e\_ihx$ and $e\_ihy$ together with the new concrete values they represent ($r\_ihx$ and $r\_ihy$) and the proofs ($p\_ihx$ and $p\_ihy$) justifying these transformation on the concrete values. Note that the simplification $[const1+const2]$ is done by calling the magma prover, which gives a new expression $e3$, even thought it could perfectly be done directly here, but we wanted to show how a prover can be reused for building another prover on an easy example. The functions of conversion between the different levels : $semiGroup\_to\_magma$ and $magma\_to\_semiGroup$ are helping for this task.
\\
A meta-variable Massoc1 has been introduced. This metavariable corresponds to a proof obligation for the following property :
$Massoc1 : Plus\ (Plus\ x\ const1)\ (Plus\ const2\ y)\ \simeq\ Plus\ (Plus\ r\_ihx\ r_3)\ r\_ihy$ in a context where there is, amongst other things, $p\_ihx : x\ \simeq\ r\_ihx$, $p\_ihy : y\ \simeq\ r\_ihy$ and $p_3 : Plus\ const1\ const2 \simeq r\_3$.
This lemma says that the treatment done by this function $assoc$ on this pattern was sound because the previous and the new concrete values are equal. It can easily be proven with a few rewritings and the use of the associativity, available from the semi-group structure. \\
\\
After this simplification of constants computable thanks to associativity, the normalization for SemiGroup also has to rearrange the brackets on a systematic way. The fully right or fully left associative forms can both be chosen for this purpose, and the implementation is very similar to the function $expr_l$ previously described for the smaller example on section 2.3.

@@

The first thing that the reduction function of the group prover will have to do is simply to transform every subtraction $(a-b)$ into $(a + (-b))$, and we are entitled to do so precisely because we have available the property $Minus\_simpl$ contained in the instance of the type class.
After that, the reduction continues with the propagation of the $Neg$ operation inside the parenthesis and $-(a+b)$ is transformed into $(-b) + (-a)$. Note that we have to be careful and not simplify it to $(-a) + (-b)$ as it would assume that we're having a commutative monoid.
Then, because we've pushed some negations inside the parenthesis, we might have sequences of two or more consecutive negations. We simplify them by removing two consecutive negations every time that we find such a sequence. Once this is done, there is a last major step specific of groups to accomplish, which is the simplification of sums of symmetric elements.

@@

\begin{figure}[H]
\figrule
\begin{center}
\begin{lstlisting}
data Variable : {c:Type} -> {n:Nat} -> (Vect n c) -> c -> Type where
    RealVariable : ($\Gamma$:Vect n c) -> (i:Fin n) -> Variable $\Gamma$ (index i $\Gamma$)
    EncodingNegOfVar : ($\Gamma$:Vect n c) -> (i:Fin n) -> Variable $\Gamma$ (Neg (index i $\Gamma$))
\end{lstlisting}
\end{center}
\caption{Real variables and encodings}
\figrule
\end{figure}

@@

The constructor for variable now takes a Variable, instead of taking directly a $(Fin\ n)$.
\begin{figure}[H]
\figrule
\begin{center}
\begin{lstlisting}
    VarG : (p:Group c) -> {$\Gamma$:Vect n c} -> {val:c} 
          -> (Variable $\Gamma$ val) -> ExprG p $\Gamma$ val
\end{lstlisting}
\end{center}
\caption{new Var constructor}
\figrule
\end{figure}

@@

The most important specific task of the ring reduction is to develop entirely the expression with the use of distributivity. After that, it needs to call the commutative group prover, but again, we're losing some expressivity as it won't be possible to express a product at the commutative group level. We follow the same idea that we've implemented for the conversion going from group to monoid, and we will use some specific encoding. More precisely, all monomials (containing products), like $3*(x*y)$ are encoded as variables when passed to the commutative group prover. This is possible because we've fully developed the expression, which had for effect to ensure that a product can't contain a sum.
We will add a coonstructor to the type $Variable$ in order to encode these monomials.

>>>>

We now describe briefly how the normalisation functions work for some of the different algebraic structures, and we also explain how we've been able to reuse each prover for the more specialised structures. The construction of the desired proof is always done bit by bit, because every rewriting of the reflected term is immediately justified by a little accompanying proof which shows the equivalence of the previous and new index.

\begin{lstlisting}
littleRewriting : (p:algebraic structure on c) $\rightarrow$ ($\Gamma$:Vect n c) 
       $\rightarrow$ {c1:c} $\rightarrow$ (Expr p $\Gamma$ c1) $\rightarrow$ (c2 ** (Expr p $\Gamma$ c2, c1 $\simeq$ c2))
littleRewriting p $\Gamma$ e = (new expression, justification)
\end{lstlisting}

@@

\begin{lstlisting}
assoc : (p:SemiGroup c) $\rightarrow$ ($\Gamma$:Vect n c) $\rightarrow$ {c1:c} 
        $\rightarrow$ (ExprSG p $\Gamma$ c1) $\rightarrow$ (c2 ** (ExprSG p $\Gamma$ c2, c1 $\simeq$ c2))
assoc p $\Gamma$ (PlusSG 
             (PlusSG ex (ConstSG _ _ const1)) 
             (PlusSG (ConstSG _ _ const2) ey)) =
    let (r_ihx ** (e_ihx, p_ihx)) = (assoc p $\Gamma$ ex) in
    let (r_ihy ** (e_ihy, p_ihy)) = (assoc p $\Gamma$ ey) in
    let (r_3 ** (e_3, p_3)) = magmaReduce (semiGroup_to_magma 
                             (PlusSG (ConstSG _ _ const1) (ConstSG _ _ const2))) in
    let e_3' = magma_to_semiGroup p e_3 in
        (_ ** ((PlusSG (PlusSG e_ihx e_3') e_ihy), ?Massoc1))
\end{lstlisting}

The treatment is applied recursively on the sub-expressions $ex$ and $ey$, which produces the new expressions $e\_ihx$ and $e\_ihy$ together with the new concrete values they represent ($r\_ihx$ and $r\_ihy$) and the proofs ($p\_ihx$ and $p\_ihy$) justifying these transformation on the concrete values. Note that the simplification $[const1+const2]$ is done by calling the magma prover, which gives a new expression $e3$, even thought it could perfectly be done directly here, but we wanted to show how a prover can be reused for building another prover on an easy example. The functions of conversion between the different levels : $semiGroup\_to\_magma$ and $magma\_to\_semiGroup$ are helping for this task.
\\
A meta-variable Massoc1 has been introduced. This metavariable corresponds to a proof obligation for the following property :
$Massoc1 : Plus\ (Plus\ x\ const1)\ (Plus\ const2\ y)\ \simeq\ Plus\ (Plus\ r\_ihx\ r_3)\ r\_ihy$ in a context where there is, amongst other things, $p\_ihx : x\ \simeq\ r\_ihx$, $p\_ihy : y\ \simeq\ r\_ihy$ and $p_3 : Plus\ const1\ const2 \simeq r\_3$.
This lemma says that the treatment done by this function $assoc$ on this pattern was sound because the previous and the new concrete values are equal. It can easily be proven with a few rewritings and the use of the associativity, available from the semi-group structure. \\
\\
After this simplification of constants computable thanks to associativity, the normalization for SemiGroup also has to rearrange the brackets on a systematic way. The fully right or fully left associative forms can both be chosen for this purpose, and the implementation is very similar to the function $expr_l$ previously described for the smaller example on section 2.3.

@@

\subsection {Reusing the monoid prover for building the group prover}

The first thing that the reduction function of the group prover has to do is to transform every subtraction $(a-b)$ into $(a + (-b))$, and we are entitled to do so precisely because we have available the property $Minus\_simpl$ contained in the instance of the typeclass.
After that, the reduction continues with the propagation of the $Neg$ operation inside the parenthesis and $-(a+b)$ is transformed into $(-b) + (-a)$. Note that we have to be careful and not simplify it to $(-a) + (-b)$ as it would assume that we're having a commutative monoid.
Then, because we've pushed some negations inside the parenthesis, we might have sequences of two or more consecutive negations. We simplify them by removing two consecutive negations every time that we find such a sequence. Once this is done, there is a last major step specific of groups to accomplish, which is the simplification of sums of symmetric elements.

@@

\begin{figure}[H]
\figrule
\begin{center}
\begin{lstlisting}
data Variable : {c:Type} -> {n:Nat} -> (Vect n c) -> c -> Type where
    RealVariable : ($\Gamma$:Vect n c) -> (i:Fin n) -> Variable $\Gamma$ (index i $\Gamma$)
    EncodingNegOfVar : ($\Gamma$:Vect n c) -> (i:Fin n) -> Variable $\Gamma$ (Neg (index i $\Gamma$))
\end{lstlisting}
\end{center}
\caption{Real variables and encodings}
\label{Variable}
\figrule
\end{figure}

@@

The constructor for variable now takes a Variable, instead of taking directly a $(Fin\ n)$.
\begin{figure}[H]
\figrule
\begin{center}
\begin{lstlisting}
    VarG : (p:Group c) -> {$\Gamma$:Vect n c} -> {val:c} 
          -> (Variable $\Gamma$ val) -> ExprG p $\Gamma$ val
\end{lstlisting}
\end{center}
\caption{new Var constructor}
\label{VarG}
\figrule
\end{figure}

@@

The most important specific task of the ring reduction is to develop entirely the expression with the use of distributivity. After that, it needs to call the commutative group prover, but again, we're losing some expressivity as it won't be possible to express a product at the commutative group level. We follow the same idea that we've implemented for the conversion going from group to monoid, and we will use some specific encoding. More precisely, all monomials (containing products), like $3*(x*y)$ are encoded as variables when passed to the commutative group prover. This is possible because we've fully developed the expression, which had for effect to ensure that a product can't contain a sum.
We add a coonstructor to the type $Variable$ in order to encode these monomials.
